Finiteness in Cubical Type Theory

### 2020-01-26 paper.tex
more

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Cardinal Finiteness and Kuratowski Finiteness}
\begin{romlemma} \label{cardinal-kuratowski}
  \begin{equation}
    \mathcal{C}(A) \simeq \mathcal{K}(A) \times \text{Discrete}(A)
  \end{equation}
\end{romlemma}
This proof is constructed by providing a pair of functions: one from
\(\mathcal{C}(A)\) to \(\mathcal{K}(A) \times \text{Discrete}(A)\), and one the
other way.
This pair implies an equivalence, because both source and target are
propositions.
The actual functions themselves are proven in our formalisation, as well as in
\cite{fruminFiniteSetsHomotopy2018}.
\section{Topos} \label{topos}
In this section we will prove that decidable Kuratowski finite types form a
\(\Pi\)-pretopos.
Along the way we will provide closure proofs for a number of the other
finiteness predicates.
As we saw in theorem~\ref{cardinal-kuratowski}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first two closure proofs, we only consider manifest Bishop finiteness:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.
\begin{romlemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are manifest Bishop finite.
\end{romlemma}
\begin{romlemma}
  Split enumerability is closed under \(\Sigma\)
\end{romlemma}
To show closure under \(\Sigma\), we take the Cartesian product of the support
lists.

\begin{romlemma}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{romlemma}
\begin{proof}
  Firstly, we will simplify things slightly by working only with split
  enumerability.
  As this is equal in strength to manifest Bishop finiteness, any closure proofs
  carry over.
  Secondly, we will replace \(A\) in all places with \(\mathbf{Fin}(n)\), using
  the equivalence provided by lemma~\ref{bishop-equiv}.

  The remainder of the proof now is concerned with proving the following:
  \begin{equation} \label{given-transp-sublemma}
    \left( \Pi {(x : \mathbf{Fin}(n))} , \mathcal{E}!(U(x)) \right) \rightarrow \mathcal{E}!\left( \Pi {(x : \mathbf{Fin}(n))} , U(x) \right)
  \end{equation}

  To prove this goal, consider the following type of \(n\)-tuples.
  \begin{equation}
    \begin{alignat}{3}
      & \mathbf{Tuple}(T, 0)   &&\coloneqq \top \\
      & \mathbf{Tuple}(T, n+1) &&\coloneqq T(0) \times \mathbf{Tuple}(T \circ \text{suc}, n)
    \end{alignat}
  \end{equation}
  We have the following equivalence, proven in our formalisation.
  \begin{equation} \label{fin-equiv-sublemma}
    \mathbf{Tuple}(T, n) \simeq \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}

  By the given proof of finiteness in equation~\ref{given-transp-sublemma}, we
  know that \(\mathbf{Tuple}(T , n)\) is split enumerable, since it is made of
  products of the points of \(U\).
  We can then transport along the equivalence in
  equation~\ref{fin-equiv-sublemma}, proving our goal.
\end{proof}

>>>>

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Cardinal Finiteness and Kuratowski Finiteness}
\begin{romlemma} \label{cardinal-kuratowski}
  \begin{equation}
    \mathcal{C}(A) \simeq \mathcal{K}(A) \times \text{Discrete}(A)
  \end{equation}
\end{romlemma}
This proof is constructed by providing a pair of functions: one from
\(\mathcal{C}(A)\) to \(\mathcal{K}(A) \times \text{Discrete}(A)\), and one the
other way.
This pair implies an equivalence, because both source and target are
propositions.
The actual functions themselves are proven in our formalisation, as well as in
\cite{fruminFiniteSetsHomotopy2018}.
\section{Topos} \label{topos}
In this section we will prove that decidable Kuratowski finite types form a
\(\Pi\)-pretopos.
Along the way we will provide closure proofs for a number of the other
finiteness predicates.
As we saw in theorem~\ref{cardinal-kuratowski}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first three closure proofs, we only consider split enumerability:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.

\begin{romlemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are split enumerable.
\end{romlemma}
\begin{romlemma} \label{split-enum-sigma}
  Split enumerability is closed under \(\Sigma\).
  \begin{equation}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi(x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma(x : A) , U(x))
    }
  \end{equation}
\end{romlemma}
From this we can derive split enumerability of non-dependent sums and products,
as both are definable in terms of \(\Sigma\).

\begin{romlemma}
  Split enumerability is closed under dependent functions.
  (\(\Pi\)-types).
  \begin{equation}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi {(x : A)} , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{romlemma}
\begin{proof}
  Let \(A\) be a split enumerable type, and \(U\) be a type family from \(A\),
  which is split enumerable over all points of \(A\).

  As \(A\) is split enumerable, we know that it is also manifestly Bishop finite
  (lemma~\ref{split-enum-to-manifest-bishop}), and consequently we know \(A
  \simeq \mathbf{Fin}(n)\), for some \(n\) (lemma~\ref{bishop-equiv}).
  We can therefore replace all occurrences of \(A\) with \(\mathbf{Fin}(n)\),
  changing our goal to:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi (x : \mathbf{Fin}(n)) , U(x)\right)
    }
  \end{equation}

  We then define the type of \(n\)-tuples over some type family \(T :
  \mathbf{Fin}(n) \rightarrow \mathbf{Type}\).
  \begin{equation}
    \begin{alignat}{3}
      & \mathbf{Tuple}(0, T)   &&\coloneqq \top \\
      & \mathbf{Tuple}(n+1, T) &&\coloneqq T(0) \times \mathbf{Tuple}(n, T \circ \text{suc})
    \end{alignat}
  \end{equation}
  We can show that this type is equivalent to functions (proven in our formalisation):
  \begin{equation}
    \Pi(x : \mathbf{Fin}(n)) , U(x) \simeq \mathbf{Tuple}(n, U)
  \end{equation}
  And therefore we can simplify again our goal to the following:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\mathbf{Tuple}(n, U)\right)
    }
  \end{equation}

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
  \qed
\end{proof}
